---
title: "Get Started with mrgsolve"
author: "Metrum Research Group"
date: 2020-10-09
output: 
  html_document:
    css: [styles/styles.css]
    number_sections: true
    highlight: pygments
    toc: true
    toc_depth: 3
---

```{r,include = FALSE}
knitr::opts_chunk$set(comment = '.')
```


# About `mrgsolve` 

- `R` package for simulation from ODE-based models
    - Free, OpenSource, GitHub, CRAN
- Language
    - Models written in `C++` inside model specification format
    - General purpose solver: `ODEPACK` / `DLSODA` (`FORTRAN`)
        - Automatically detect and switch between non-stiff (Adams) and stiff (BDF)
          methods for solving the differential equations
    - Simulation workflow in `R`
- Hierarchical (population) simulation
    - `ID`, $\eta$, $\varepsilon$
- Integrated PK functionaility
    - Bolus, infusion, `F`, `ALAG`, `SS` etc, handled under the hood
    - 1- and 2-cmt PK models in closed-form
- Extensible using `R`, `C++`, `Rcpp`, `boost`, `RcppArmadillo`
- `R` is it's natural habitat

## Background

- Motivation: large bone/mineral homeostatsis model (CaBone)
- History using
    - Berkeley Madonna
    - WinBUGS
    - NONMEM (attempted)
- 2010: write `R` front end to `deSolve`
- 2012: write `C++` interface to `DLSODA`
- Develop dosing / event capability
- More recently, expose functionality provided by
    - `Rcpp` - vectors, matrices, functions, environments, random numbers 
    - `boost` - numerical tools in `C++`
    - users' own `C++` code (functions, data structures, classes)
- Translator from `SBML` to `mrgsolve` using `R` bindings to `libSBML`   


## Orientation

- https://CRAN.R-project.org/package=mrgsolve

- GitHub site: https://github.com/metrumresearchgroup/mrgsolve

- Issues and questions: https://github.com/metrumresearchgroup/mrgsolve/issues

- mrgsolve website: https://mrgsolve.github.io

- User Guide: https://mrgsolve.github.io/user_guide

- Blog: https://mrgsolve.github.io/blog

- Vignettes: https://mrgsolve.github.io/vignettes

- Compare against NONMEM: https://github.com/mrgsolve/nmtests

## What we will cover today

1. Three basic workflows
1. Loading the model into R
1. Event objects
1. Data sets

Emphasis is on getting you running your own simulations today.


# Three (basic) simulation workflows

```{r setup}
library(tidyverse)
library(mrgsolve)
```

- Single profile
- Batch
- Population


These aren't entirely different, but I like to organize this way. When I'm 
planning an simulation, I first think "what type of output do I want?" and the
answer to that question directs me on what to do next.

## Single profile


This is how we load a simulation model into mrgsolve.  

Load a two-compartment model from the internal library

```{r}
mod <- modlib("pk2")
```

We now have a 2-compartment PK model with which we can simulate. It is important
to know how this works and we will talk in depth about this. But for now, let's
simulate some stuff. 

First, we'll just simulate from this model object (`mrgsim()`)

```{r}
mrgsim(mod)
```


In the output

- Essentially a data frame of simulated data
  - First column `ID`
  - Second column: `time`
  - Next columns: compartments
  - Last columns: derived quantities

___Investigate the model object a bit___

- overview

```{r}

```

- parameters


```{r}

```

- compartments

```{r}

```

- outputs

```{r}

```


### Event object

Now, we'll create an "event object" to simulate from.  This is just a concise
statement of some intervention. Like a one-liner ... easy to make. 

Let's do 100 mg x1 to the first compartment, then simulate:

```{r}
mod %>% ev(amt = 100) %>% mrgsim()
```


We use `ev()` to create a set of intervention(s) for the simulation.  Here, it
is just a single 100 mg dose into the first compartment.  The event object 
looks like this:

```{r}
ev(amt = 100)
```

We have the following columns

1. `time` - whatever is your model time
1. `amt` - whatever is the mass unit for your compartments
1. `cmt` could be number or name
1. `evid` just like nonmem - mostly using 1

You can also use:   
- `rate` - infusion   
- `ss` - steady state (1 or 2)  
- `ii` - interdose interval   
- `addl` - additional doses   
- `tinf`  - infusion time (rather than `rate`)  
- `total` - total number of doses (rather than `addl`)  

See `?ev`

### Plot

Simulate 100 mg x1 again and now we pipe it to `plot()`

```{r}
mod %>% ev(amt = 100) %>% mrgsim() %>% plot()
```

### Control time span of the simulation 

I would like this to look a little nicer. 

- 100 mg x1    
- Run the end of the simulation out to 72 hours with delta 0.1 
- Make the line smoother  
- Plot the result  

```{r}
mod %>% ev(amt = 100) %>% mrgsim(end = 72, delta = 0.1) %>% plot()
```

We can make this change permanent

- end: 72 hours
- delta: 0.1 hours

```{r}
mod2 <- update(mod, end = 72, delta = 0.1)
```



### More-complicated events

We said that the `event` objects were simple.  But we can combine them to make 
more complicated sequences.  

Let's load a PK model for azithromycin (`azithro-single`):

```{r}
mod <- mread("azithro-single", project = "model")
```

__Check out the model__

```{r}
mod
```

Create an event object to implement the z-pak dose:

- 500 mg po on day 1 (`load`)
- 250 mg po daily on days 2 through 5 (`continue`)

```{r}
load <- ev(amt = 500)
continue <- ev(amt = 250, ii = 24, addl = 3, time = 24)
zpak <- c(load, continue)
```


Look at the zpak dosing object

```{r}
zpak
```


We can also accompilsh this just with 250 mg tablets

```{r}
zpak <- c(ev(amt = 250), ev(amt = 250, ii = 24, addl = 4))

zpak
```

Now, simulate and plot from the zpak event object

```{r}
mrgsim(mod, zpak) %>% plot()
```


### Event sequence

- 100 mg daily x 7 __then__   
- 50 mg BID x7   

```{r}

```



## Batch

Let's use our fixed-effects azithromycin model to look at how weight affects
PK. We'll use that `zpak` object that we created in the previous section. 

### Create an idata set

Now, let's make a data frame that contains the weights that we want to
investigate (from 40 kg to 140 kg by 10 kg). 

```{r}
wt <- tibble(WT = seq(40, 140, 10))

head(wt)
```


__IMPORTANT__: the key here is that we have `WT` as a column in the data set
and we have `WT` as a parameter in the model (look at the parameters)

```{r}
param(mod)
```

When we make the names agree, mrgsolve will update the `WT` parameter as the 
simulation advances across individuals.


### Simulate with event object

Now we can pass this set of weights into the problem as "idata".  We will use
just the first record from the zpak dosing object.

- Load the `azithro-single` model
- Create a dosing event with 500 mg x1
- Simulate with `idata`
- End the simulation at 96 hours

```{r}
mod <- mread("azithro-single", project = "model")

load <- ev(amt = 500)

out <- mrgsim(mod, events = load, idata = wt, end = 96)
```

Take a quick look at the output (`head`)
```{r}
out
```



Plot the ouptut, looking only at `CP` and on log scale

```{r}
plot(out, CP ~ time,  logy = TRUE)
```


This idata set functionality is typically used with an event object 
(as we have here) but isn't required.


## Population

The last workflow I'm calling "population".  Here, population just refers to 
the input data set. 

We can have a data frame that contains many individuals with all different 
types of dosing interventions.  This is just like the data set that you use
to do your NONMEM run. 

### Read a NMTRAN like data set

Meropenem PopPK
http://repository.ddmore.foundation/model/DDMODEL00000213


For example, read in `data/meropenem.csv`

```{r}
data <- readr::read_csv("data/meropenem.csv", na = '.')
```


- glimpse the data (`head`)
- count `EVID`, `DUR`, `AMT`
- number of IDs

```{r}
head(data)

count(data, EVID, DUR, AMT)

length(unique(data$ID))
```


Now, load the meropenem model (`meropenem_pk.cpp`, in the `model` directory)
```{r}
mod <- mread("meropenem_pk", project = "model")

```

And simulate with `mod` and `data`
```{r}
out <- mrgsim(mod, data)
```


Then plot using `Y` output

```{r}
plot(out, Y ~ TIME)
```


Resimulate and plot by duration

```{r}
out <- mrgsim(mod, data, carry_out = "DUR")
```

Plot `Y` versus `TIME` by `DUR`
```{r}
plot(out, Y ~ TIME | DUR)
```

Recall that we have both observations and doses in the data set (as usual 
for your NONMEM data set)

Count `EVID` in `data`

```{r}
count(data, EVID)
```


When mrgsolve finds records with `EVID=0` in the data set, it will assume that 
you have specified every time that you want a simulated value in the data set.
In other words the design of the simulated output will match the design of the 
input data:


Check `dim()` in `data` and `out`:

```{r}
dim(data) 
dim(out)
```


Let's look to see what happens when we don't include any observation records
in the input data:

Filter into `doses`:

```{r}
doses <- filter(data, EVID==1)
```

Now we still have the same number of people, with different doses and infusion
times. 

Check unique `ID` and count `EVID`, `AMT`, and `DUR`

```{r}
length(unique(doses$ID))
count(doses, EVID, AMT, DUR)
```


- Simulate from this sparse data set; end = 8 hours
- Get `DUR` in to the output
- Plot `log(Y)` versus `time` by `DUR`


```{r} 
mod %>% 
  mrgsim(doses, carry_out = "DUR", end = 8) %>% 
  plot(Y~TIME|factor(DUR), logy=TRUE)
```


The principle is: when mrgsolve does NOT find any observation records, it will 
fill them in for you according to the time grid that we looked at previously. 

This can be very helpful in reducing the data assembly burden when running your 
simulations.


### Some other ways to create population inputs


- `expand.ev()` makes all combinations of your inputs
  - Sensible defaults are provided
  - `time`, `cmt`, `evid`

```{r}
data <- expand.ev(amt = c(100, 300, 1000), ii = c(12, 24))

data
```


- `as_data_set()` takes event objects and creates a data frame / set
  - 100 mg and 300 mg doses daily x7
  - 20 individuals each

```{r}
data <- as_data_set(
  ev(amt = 100, ii = 24, total = 7, ID = 1:20), 
  ev(amt = 300, ii = 24, total = 7, ID = 1:20)
)

```


# Get work done


## Work with output

- names
- summary
- head
- `$`

```{r}
mod <- modlib("pk1")

out <- mrgsim(mod)

names(out)
summary(out)
head(out)
out$time[1:5]
```



## Coerce output

- data.frame
- tibble
- matrix

```{r}
out <- mrgsim(mod)

class(out)

as_tibble(out)

```

## Corerce via dplyr verbs

- Simulate and pipe the output to `mutate()`

```{r}
out <- mrgsim(mod) %>% mutate(name = "kyle")

class(out)
head(out)
```



## Return data frame

- Use `output` argument

```{r}
out <- mrgsim(mod, output = "df")

class(out)
```

- Use `mrgsim_df()`

```{r}
out <- mrgsim_df(mod)

class(out)
```


## Carry-Out

- 100 mg x1
- need `dose` in the output
- contrast that with getting `amt` in the output

First create the event
```{r}
e <- ev(amt = 100, dose = amt)

e
```

Then recover `dose`
```{r}
mrgsim(mod, events = e, carry_out = "dose")
```


## Recover

- `dose`: 100 mg
- `trt`: 100 mg x1
- need `dose` and `trt` in the output

First, create the event
```{r}
e <- ev(amt = 100, trt = "100 mg x1", dose = amt)

e
```

Then simulate and recover `trt` and `amt`
```{r}
mrgsim(mod, events = e, recover = "trt,amt")
```


