--- 
title: "Pooled analysis of indomethacin PK data"
author: Metrum Research Group
date: ""
---

# Packages

```{r,echo=FALSE}
knitr::opts_chunk$set(comment='.', message=FALSE, fig.path="figures/indo-", 
                      cache=TRUE, autodep=TRUE)
```

```{r}
library(tidyverse)
theme_set(theme_bw())
library(mrgsolve)
```

# Load indomethacin data set
```{r}
data(Indometh)
```

- Take a look at what is there
```{r}
head(Indometh)

count(Indometh, Subject)

ggplot(Indometh, aes(time,conc,group=Subject)) + 
  geom_point() + geom_line() +
  scale_y_continuous(trans = "log", breaks = 10^seq(-4,4))
```


This is individual-level data, but we are going to do naive pooled 
analysis. 

# Data assembly 
```{r}
data <- readRDS("data/indometh.RDS")

head(data)
```

# Load a PK model 

- We'll try out one-compartment first
```{r}
mod <- modlib("pk1")

param(mod)
```

Pick some parameters to estimate:
```{r}
theta <- log(c(CL = 1, V = 100))
```

```{r}
names(theta)
```


# Create an objective function function

- For starters, just do OLS estimation
- Note that we _need_ to name the parameters (`p`)
    - Parameter updates require names in `mrgsolve`
    - Generally, don't expect `p` to retain any names that you might
    pass in through the initial estimates
- We also pass in the `data` and the dependent variable (`dv`)

```{r}
obj <- function(p, theta, data, dv ="conc", pred = FALSE) {
  
  names(p) <- names(theta)
  
  p <- lapply(p,exp)
  
  mod <- param(mod, p)
  
  out <- mrgsim_q(mod, data, output="df")
  
  if(pred) return(out)
  
  sqr <- (out[["CP"]] - data[[dv]])^2
  
  sum(sqr, na.rm=TRUE)
}
```



# Fit with one-compartment model

- First generate some initial estimates
- These _need_ to be named in a way that is consistent with
the model we are using
- I usually run a test with the objective function function to 
make sure the logic works out

```{r}
obj(theta,theta,data)
```

- Nelder-Mead optimization
```{r}
fit <- optim(par = theta, fn=obj, theta = theta, data=data)
```

- And generate some predictions based on the final estimates
```{r}
pred <- obj(fit$par, theta, data, pred = TRUE)

data$pred <- pred$CP

head(data)
```

# Make a plot of the output

- What do you think?  Good fit?

```{r, warning = FALSE}
ggplot(data = data) + 
  geom_point(aes(time,conc)) + 
  scale_y_log10() + 
  geom_line(aes(time,pred),col="firebrick", lwd=1)
```

# Your turn

- Try fitting the same indomethacin data with a 2-compartment model

```{r}
mod <- modlib("pk2")
```

- Take a look at the model and generate a call to `minqa::newuoa` using 
the OLS objective function above to fit the data

- You will also need try out a new set of initial estimates for all of
the volumes and clearances for 2-compartment, IV bolus model

- What do you think of the fit using the the OLS objective function?
    - Can you make a simple modification to the OLS objective function
    that might make the fit look a little better?

- Suppose we're worried about the `newuoa` optimizer and want to 
try a global search algorithm
    - Can you construct a call to `RcppDE::DEoptim` that will also 
    fit the data?
    - Remember that `DEoptim` doesn't use initial estimates the 
    same way `stats::optim` or `minqa::newuoa` does; you have to 
    specify one vector of lower boundaries and one vector of 
    upper boundaries, with a lower and upper bound for each parameter



# Answer

- Set the initial estimates for two compartment model

```{r}
param(mod)

theta <- log(c(CL = 2, V2 = 50, Q = 10, V3 = 50))
```


```{r}
fit <- optim(par = theta, fn=obj, theta = theta, data=data)
```

- And generate some predictions based on the final estimates
```{r}
pred <- obj(fit$par, theta, data, pred = TRUE)

data$pred <- pred$CP

ggplot(data = data) + 
  geom_point(aes(time,conc)) + 
  scale_y_log10() + 
  geom_line(aes(time,pred),col="firebrick", lwd=1)
```


- Try weighted least squares 

```{r}
obj <- function(p, theta, data, wt, pred = FALSE) {
  names(p) <- names(theta)
  p <- lapply(p,exp)
  out <- mod %>% param(p) %>% mrgsim_q(data, output="df")
  if(pred) return(out)
  return(sum(((out$CP - data[["conc"]])*wt)^2, na.rm=TRUE))
}
```

```{r}
dv <- data[["conc"]]

fit_wt <- minqa::newuoa(par = theta, fn=obj, theta = theta, data=data, wt=1/dv)
```

Final estimates and final value of objective function
```{r}
exp(fit_wt$par)

obj(fit_wt$par,theta,data,dv)
```

- Generate predictions for the final and initial estimates

```{r}
pred <-  obj(fit$par, theta, data, wt = 1/dv, pred = TRUE)
predi <- obj(theta,  theta, data, wt = 1/dv, pred = TRUE)
predw <- obj(fit_wt$par, theta, data, wt = 1/dv, pred = TRUE) 


data$pred <- pred$CP
data$predi <- predi$CP
data$predw <- predw$CP
head(data)
```

- Plot the predictions
```{r}
pred <- distinct(data, time, .keep_all = TRUE)

ggplot(data = data) + 
  geom_point(aes(time,conc)) + 
  scale_y_log10() + 
  geom_line(data=pred,aes(time,pred),col="black", lwd=1, alpha = 0.6) +
  geom_line(data=pred,aes(time,predi),col="darkgreen", lwd=1) + 
  geom_line(data = pred, aes(time,predw), col="firebrick", lwd = 1)
```

## Fit the data with `RcppDE::DEoptim`

```{r, eval = FALSE}
fit <- DEoptim::DEoptim(
  obj, 
  lower = rep(-4,4), 
  upper = rep(4,4), 
  theta = theta, data = data, wt = 1/dv, 
  control = DEoptim::DEoptim.control(itermax=120,trace=20)
)

```

## Check the estimates and the final value of the objective function

```{r, eval = FALSE}
tibble(
  DE = exp(fit$optim$bestmem), 
  Nelder = exp(fit_wt$par)
)

tibble(
  DE = obj(fit$optim$bestmem, theta,data,1/dv),
  Nelder = obj(fit_wt$par, theta, data, 1/dv)
)
```



# Some global search with NLOPTR

https://nlopt.readthedocs.io/en/latest/NLopt_Algorithms/

```{r, eval=FALSE}
library(nloptr)
a0 <- obj(theta,theta=theta,data=data,wt = 1/dv, pred = FALSE)

lowr <- rep(-5,length(theta))
uppr <- rep(5, length(theta))

x <- isres(
  x0 = theta, 
  fn=obj, 
  lower = lowr, 
  upper = uppr, 
  theta=theta, 
  data=data, 
  wt = 1/dv, 
  maxeval=20000
)

y <- crs2lm(
  x0 = theta, 
  fn=obj, 
  lower = lowr, 
  upper = uppr, 
  theta=theta, 
  data=data, 
  wt = 1/dv, 
  maxeval=5000
)

z <- newuoa(x0 = y$par, fn = obj,theta = theta, data = data, wt = 1/dv)


tibble(a0 = theta, a = fit_wt$par, x = x$par, y = y$par, z= z$par) %>% exp
tibble(a0 = a0,a = fit_wt$fval, x = x$value, y= y$value, z = z$value)


direct <- directL( 
  fn = obj, 
  lower = lowr, 
  upper = uppr, 
  theta = theta, 
  data = data, 
  wt = 1/dv, 
  control = list(maxeval=2500)
)


tibble(a0 = theta, a = fit_wt$par, x = x$par, y = y$par, z= z$par,d = direct$par) %>% exp
tibble(a0 = a0,a = fit_wt$fval, x = x$value, y= y$value, z = z$value, d = direct$value)

```



