---
title: Introduction to parameter optimization
author: "Metrum Research Group"
date: ""
---

```{r, setup,  message = FALSE}
library(tidyverse)
library(broom)
library(ggplot2)
theme_set(theme_bw())
```


```{r, setup2, echo = FALSE, message = FALSE}
source("script/global.R")
knitr::opts_chunk$set(fig.path = "figures/intro-")
options(mrgsolve.soloc="build", dplyr.summarise.inform = FALSE)
```


# Swiss Fertility and Socioeconomic Indicators

We'll get started with the introduciton to optimization with an easy 
data set and an easy model. 

Data were collected from 47 French-speaking provences around 1888. 
```{r}
data(swiss)
glimpse(swiss, width = 60, strict.width="cut")
```

```{r}
ggplot(swiss, aes(Examination,Fertility)) + geom_point() + geom_smooth()
```

We'll work on a regression model for the standardized fertility 
measure (dependent variable) as function of the `Examination`
predictor (percent draftees receiving highest mark on 
army examination).


Usually we'd fit this model in R like this:
```{r}
fit <- lm(Fertility ~ Examination, swiss) 

tidy(fit)
```


Using `lm` is the "right" way to model this data.  We're going to write 
come code that also get parameters for this data.  But we'll put all of the 
pieces together ourselves.  So this isn't the proper way to get these parameter
estimates.  But this simple example will help us better understand the 
mechanics of parameter optimization in R.

# Optimization the `R` way

We'll need

  1. A __model__
      - The model generates the data (dependent variable) based on parameters and 
      predictors
  1. Some __data__
      - Using `swiss` for now
  1. An __objective function__
       - To quantify how consistent a set of parameters are with the 
      observed data
  1. An __optimizer__
      - Search the parameter space for most optimal parameter value

## Model

The parameters are 

- `intercept`
- `slope`

And the predictor is `ex` ... in our example `Examination` 

This is pretty simple stuff.  But we'll wrap it up in a function to 
call like this:

```{r}
linear_model <- function(intercept, slope, ex) {
  intercept + slope * ex
}
```

So we can get the predicted `Fertility` by passing in the intercept, 
slope, and the `Examination` value

E(y|x)
```{r}
linear_model(intercept = 90, slope = -2, ex = 20)
```

## Data

How you want to set this up is a bit dependent on your application.  I'm 
going to get vectors called `ex` for the `Examination` value (predictor)
and `fer` for `Fertility` (the dependent variable).

```{r}
ex <- swiss[["Examination"]]

fer <- swiss[["Fertility"]]
```

## Objective function

We'll write this function so that 

  1. The first argument is `par`, the parameters we want to evaluate
      - Par will be a vector of length 2, the intercept and the slope
  1. We will also pass in the predictor (`ex`) and the data (`fer`), which 
     we'll need to calculate the objective function


```{r}
ofv <- function(par, ex, fer) {
  
  fer_hat <- linear_model(par[1], par[2], ex)
  
  sum((fer-fer_hat)^2)
}
```

This is an _O_ rdinary _L_ east _S_ quares objective function.  

Working backward:

1. We return the squared difference between the predicted values (`fer_hat`) and 
the data
1. We generate the predicted values from our linear model function, 
the proposed parameters and the data
1. The optimizer will propose a set of parameters for us to evaluate


Let's test the objective function

```{r}
theta <- c(70, -2)

ofv(theta, ex, fer)
```

Good or bad?  Looking back at the data, the intercept doesn't look like 
it is 70 ... more like 80.  Let's try that:

```{r}
theta <- c(80,-2)

ofv(theta, ex, fer)
```

Ok the objective function is lower now.  The second set of parameters we tried
looks better than the first set.  

What about slope?

```{r}
theta <- c(80,-1.5)

ofv(theta, ex, fer)
```

This is even better.  But we can't keep going like this.  

## Parameter search

Let's do this for a big batch of parameters

- intercept from 75 to 95
- slope from -2 to 0

```{r}
test <- expand.grid(intercept = seq(75,95,1), slope = seq(-2,0,0.1))

head(test)
```

Now calculate the value of the objective function for each paramter set
```{r}
test <- mutate(
  test, 
  value = pmap_dbl(test, .f=function(intercept,slope) {
    ofv(c(intercept,slope), ex = ex, fer = fer)
  })
)

arrange(test,value) %>% head

ggplot(test) + geom_contour(aes(intercept,slope,z=value),bins=80)
```



## Optimize

We know there is a set of parameters that really gets us the smallest
value of the objective function and are therefor the "optimal" parameters.

We invoke an optizer in R to search the parameter space and find that set of 
parameters.

Start with an optimizer that comes with R in the `stats` package.  `optim`
by default does Nelder-Mead optimization algorithm.

When we call `optim`, we have to give an inital guess (`par`) and 
the function to minimize (`ofv`).  We also pass in the 
predictor and the vector of observed data so we can calculate the 
sum of squares.

```{r}
fit <- optim(c(100,1), ofv, ex = ex, fer = fer)
```



```{r}
fit$par

lm(Fertility~Examination, swiss) 
```


# Use other optimizers

## `minqa::newuoa`

```{r}
library(minqa)

fit <- newuoa(theta, ofv, ex = ex, fer = fer, control = list(iprint=20))
```

```{r}
fit$par
```

## DEoptim

Differential evolution algorithm

```{r, output.size = 160}
library(DEoptim)

lower <- c(intercept=0, slope=-100)
upper <- c(intercept = 1000, slope=100)

con <- DEoptim.control(itermax = 80, trace = 2)

set.seed(112233)
fit <- DEoptim(ofv, lower, upper, ex = ex, fer = fer, control = con)
```


# Maximum likelihood estimation

Let's write a new (R) function where we optimize based on a normal likelihood
function. 

The arguments are the same as the OLS function.  Now, rather than comparing 
predictions against data using sum of squares, we compare based on 
normal likelihood function.


```{r}
ml <- function(p, ex, fer) {
  
  fer_hat <- linear_model(p[1], p[2], ex)
  
  like <- dnorm(fer, fer_hat, p[3], log = TRUE)
  
  -1*sum(like)
}
```


__Note__

1. We have an extra parameter now ... the standard deviation for likelihood
function
1. We use `log=TRUE` to get the log likelihood; then the joint likelihood 
of all the data is the sum of the individual likelihoods
1. We return minus-1 times the log likelihood; we are doing *maximum* likelihood
but the optimizers find the *minimum* of a function


Test the function now

```{r}
theta <- c(intercept = 10, slope = 1, sd = 2)

ml(theta, ex, fer)
```

And we get the same answer
```{r}
fit <- newuoa(theta, ml, ex = ex, fer = fer)

fit$par

fit$fval
```

# Extended least squares

```{r}
els <- function(p, ex, fer) {
  
  fer_hat <- linear_model(p[1], p[2], ex)
  
  0.5 * sum((fer - fer_hat)^2/p[3] + log(p[3]))
}
```

```{r}
fit.els <- newuoa(theta, els, ex = ex, fer = fer)

fit.els$par

fit.els$fval
```


# Get standard error of estimate

We use `numDeriv::hessian` to get the hessian

```{r}
library(numDeriv)

he <- hessian(ml, fit$par, ex = ex, fer = fer)

he
```


To derive the standard error

1. Invert the hessian matrix
1. Get the diagonal elements
1. Take the squre root

```{r}
he %>% solve() %>% diag() %>% sqrt()
```

And compare against the answer we got from `lm`
```{r}
lm(Fertility ~ Examination, data = swiss) %>% 
  tidy() %>% 
  pull(std.error)
```


You can also try `nlme::fdHess` 
```{r}
library(nlme)

he <- fdHess(pars  = fit$par, fun = ml, ex = ex, fer = fer)

he$Hessian %>% solve() %>% diag() %>% sqrt()
```

In my experience, it is frequently necessary to just bootstrap the data
set.  We will look at likelihood profile in a separate vignette.

# Plot predicted and observed values

Take the final parameter estimates
```{r}
fit$par
```

and pass them into our `linear_model` to generate predicted values.


```{r}
data <- tibble(
  ex = ex, 
  fer = fer, 
  pred = linear_model(fit$par[1], fit$par[2], ex)
)
```

```{r}
ggplot(data = data) + 
  geom_point(aes(x =  ex, y = fer)) +
  geom_line(aes(x = ex, y = pred), lwd = 2, col="red3") 
```



# Let's try it

```{r}
data <- readRDS("data/pdfit.RDS")

head(data)
```

```{r}
ggplot(data, aes(auc,response)) + geom_point()
```


