---
title: Introduction to parameter optimization
author: "Metrum Research Group"
date: ""
---


```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(mrgsolve)
library(FME)
options(mrgsolve.soloc = "build")
```


```{r, setup, echo = FALSE, message = FALSE}
source("script/global.R")
knitr::opts_chunk$set(fig.path = "figures/sensitivity_local-", 
                      cache=TRUE, autodep=TRUE)
```


# Sensitivity analysis with PBPK model

- For a small change in a model parameter `p`, what is the change in model 
output `y`?
- __Local__ sensitivity analysis
- Use the `sensFun` function from the FME package


## Load the model

```{r, warning = FALSE, message = FALSE}
mod <- mread_cache("model/yoshikado.cpp")
mod <- update(mod, end = 12, delta = 0.025, atol=1E-12,rtol=1E-12)
```

## Load a data set

```{r}
data <- read_csv("data/fig4a.csv") 
data <- mutate(data, DV = ifelse(DV < 0, NA_real_, DV))
data <- filter(data, ID==2)
dose <- filter(data, evid==1)
```

Statin / cyclosporine DDI
```{r}
mod %>% mrgsim_d(dose,delta=0.025) %>% plot(CP~time)
```

## Define a function for sensitivity analysis

Do what we just did, but wrap it up in a function and pass in some parameters.

```{r}
fun <- function(pars,data) {
  mod %>% 
    param(pars) %>%
    mrgsim_d(dose,obsonly=TRUE,output="df") %>%
    select(-ID)
}
```

Just like in the optimization function, we update the model object with 
whatever parameters were passed in and simulate.  

IMPORTANT to return a data frame of simulated data

## Pick parameters for sensitivity analysis

These are the parameters that we were focusing on in the regression model.  Adding
`Vadi` here as a negative control. 

```{r}
pars <- as.numeric(param(mod))[c("fbCLintall", "ikiu", "fbile", "ka", "ktr", "Vadi")]

pars
```


Call `sensFun` from the FME package

  - `func` - the sensitivity function we defined above
  - `parms` - parameters to investigate
  - `sensvar` - the output(s) that you want to look at
  - `tiny` the step size for sensitivity analysis
  - `data` this is an argument for our sensivitity function

```{r}
locSens <- FME::sensFun(
  func=fun, 
  parms=pars, 
  sensvar="CP", 
  tiny=1e-5, 
  data=data
)
```

## Summarize
```{r}
summary(locSens)
```

__Summary plots__

```{r}
plot(locSens, legpos="topright", lwd=2)

plot(summary(locSens))
```

A nicer view

```{r}

summ <- 
  as_tibble(summary(locSens)) %>%
  mutate(parms = names(pars)) 

ggplot(data=summ, aes(x=reorder(parms, Mean), y=Mean)) + 
  geom_col() + 
  labs(x="Parameter", y="Coefficient") +
  coord_flip() +
  geom_hline(yintercept = 0, lty=2) 
```

```{r}
#nicer view
df_temp <- as_tibble(locSens) %>%
  gather(Parameter, Coefficient, -x, -var) %>%
  mutate(Parameter = factor(Parameter)) %>%
  rename(time=x) %>%
  group_by(Parameter) %>%
  mutate(Coefficient = Coefficient - first(Coefficient)) %>%
  ungroup()

ggplot(data=df_temp, aes(x=time, y=Coefficient, col=Parameter)) +
  geom_line(lwd=1) +
  theme(legend.position="right") +
  facet_wrap(~var)

```



```{r}
fun2 <- function(pars,data) {
  mod <- param(mod, pars)
  out <- mrgsim_d(mod,data,output="df")
  wres <- (out[["CP"]] - data[["DV"]]) * 1/data[["DV"]]
  ofv <- sum(wres, na.rm=TRUE)
  tibble(time=12,ofv  = ofv)
}


locSens <- sensFun(func=fun2, parms=pars, tiny=1e-5,data=data)

summ <- 
  as_tibble(summary(locSens)) %>%
  mutate(parms = names(pars)) 

ggplot(data=summ, aes(x=reorder(parms, Mean), y=Mean)) + 
  geom_col() + 
  labs(x="Parameter", y="Coefficient") +
  coord_flip() +
  geom_hline(yintercept = 0, lty=2) 
```

