---
title: "A workflow in R for PK and PBPK modeling: simulation, optimization, and visualization"
author: R/Pharma 2020 Workshop
date: <blu>Kyle Baron & Devin Pastoor / MetrumRG</blu>
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE)
options(mrgsolve.soloc = "build")
```


```{r, include = FALSE}
library(mrgsolve)
library(dplyr)
library(knitr)
library(lattice)

```


# Optimization in R with mrgsolve

<hr>

https://github.com/metrumresearchgroup/r-pharma-pkpd-2020

<hr>

- mrgsolve intro
- The R "way" to do optimization (simple linear model)
- Indomethacin PK model optimization
- Pitavastatin / CsA DDI model optimization
- Look at options for different optimizers
- Local sensitivity analysis
- Likelihood profile
- Generate virtual population

# Single level of random effects

- Today, optimizing parameters for models with single level
of random effects (no IIV)
- Options are available: `nlme`, `saemix`
    - Use if really needed (like  [here](https://metrumrg.com/wp-content/uploads/2017/10/Baron_ASBMR_2013.pdf))
- Look at [nlmixr](https://github.com/nlmixrdevelopment/nlmixr) for multi-level modeling in R with ODEs


# About `mrgsolve` 

- `R` package for simulation from ODE-based models
    - Free, OpenSource, GitHub, CRAN
- Language
    - Models written in `C++` inside model specification format
    - General purpose solver: `ODEPACK` / `DLSODA` (`FORTRAN`)
        - Automatically detect and switch between non-stiff (Adams) and stiff (BDF)
          methods for solving the differential equations
    - Simulation workflow in `R`
- Hierarchical (population) simulation
    - `ID`, $\eta$, $\varepsilon$
- Integrated PK functionaility
    - Bolus, infusion, `F`, `ALAG`, `SS` etc, handled under the hood
    - 1- and 2-cmt PK models in closed-form
- Extensible using `R`, `C++`, `Rcpp`, `boost`, `RcppArmadillo`
- `R` is it's natural habitat

# `mrgsolve` started as `QSP` modeling tool
- Motivation: large bone/mineral homeostatsis model (CaBone)
- History using
    - Berkeley Madonna
    - WinBUGS
    - NONMEM (attempted)
- 2010: write `R` front end to `deSolve`
- 2012: write `C++` interface to `DLSODA`
- Develop dosing / event capability
- More recently, expose functionality provided by
    - `Rcpp` - vectors, matrices, functions, environments, random numbers 
    - `boost` - numerical tools in `C++`
    - users' own `C++` code (functions, data structures, classes)
- Translator from `SBML` to `mrgsolve` using `R` bindings to `libSBML`   


# Orientation

- https://CRAN.R-project.org/package=mrgsolve

- GitHub site: https://github.com/metrumresearchgroup/mrgsolve

- Issues and questions: https://github.com/metrumresearchgroup/mrgsolve/issues

- mrgsolve website: https://mrgsolve.github.io

- User Guide: https://mrgsolve.github.io/user_guide

- Blog: https://mrgsolve.github.io/blog

- Vignettes: https://mrgsolve.github.io/vignettes


# Optimization - log likelihood function

<BR><BR><BR>
<img src = "../images/setup1.png" width = "1024px"></img>

# Optimization - log likelihood function


<img src = "../images/setup2.png" width = "1024px"></img>

# Optimization - log likelihood function


<img src = "../images/setup2.5.png" width = "1024px"></img>


# Optimization - log likelihood function


<img src = "../images/setup3.png" width = "1024px"></img>


# Optimization - log likelihood function

<img src = "../images/setup4.png" width = "1024px"></img>


# Many options for optimization in R

CRAN Task View: https://cran.r-project.org/web/views/Optimization.html





